---
title: Matching and Mismatching Lead Isotope Ratios 
author: 
  - name: Yiu-Kang Hsu
    orcid: 0000-0002-2439-4863
    corresponding: true
    email: Yiu-Kang.Hsu@bergbaumuseum.de
---

## Learning objective

::: {style="text-align: justify"}
In this chapter, we will employ various statistical methods to compare lead isotope ratios between archaeological samples and reference ore data. In principle, provenance studies of lead isotopes aim to exclude unlikely ore signatures to narrow down potential sources. This "exclusion exercise" often involves defining ore fields, discriminating between possible ore sources, and estimating the influence of recycling and mixing when provenance inquiries are inconclusive.
:::

## Prior knowledge

::: {style="text-align: justify"}
It is assumed that readers are familiar with concepts of the previous chapter concerning data visualisation. Basic understanding of R is desired but not required as the proposed statistical methods can also be realised through other programming languages(e.g., python and MATLAB)
:::

## Outline

::: {style="text-align: justify"}
This section introduces various statistical concepts and techniques recently applied to the study of lead isotopes in archaeological metals. We utilise a suite of R packages to showcase how these methods can be applied to a dataset of Roman lead objects published by @Bode.2009.
:::

The following statistical methods are discussed:

<ul>

<li>Euclidean distance</li>

<li>Analytical uncertainties</li>

<li>Dimensional reduction and cluster analysis</li>

<li>Kernel density estimation</li>

<li>Other recent developments</li>

</ul>

## Learning content

The lead isotope method primarily serves as an "exclusion" technique, helping to rule out possible geological origins based on available reference data. This is necessary because ore deposits or districts from different locations may have overlapping isotopic signatures. Identifying a match does not conclusively determine the provenance, as multiple sources might share a similar range isotopic ratios.

Traditionally, an "eyeballing" method has been used, where isotopic ratios are visually compared through bivariate scatter plots, typically using 204Pb or 206Pb as the denominator. Researchers would conclude on provenance by assessing the similarity or dissimilarity between artefacts and ore samples. However, this approach can be unreliable and subjective in some circumstances, especially with large, complex databases that contain overlapping ratios from multiple potential ore sources.

Applying statistical methods to isotope data interpretation offers researchers an alternative perspective, enabling them to recognise patterns within the data and, in some cases, distinguish one source from another. Lead isotope data may also display mixing patterns, where two or more sources contribute to the final isotopic ratios. To address this, some mixing models have been proposed to help understand the life cycle of archaeological metals.

### Euclidean distance

Let's start with loading the required R packages:

```{r}
#| label: load-pkgs
#| code-summary: "Packages"
#| message: false
#| warning: false

library(plotly) # for interactive plots
library(ggtern) # for ternary kernel density plots
library(downlit)# code linking
library(xml2)   # code linking
library(dplyr)  # for tiding up data sets
library(sfsmisc) # kde probabilities 
library(fields)  # calculation of distance between points
library(kableExtra) # for coloring fields of a table
```

Euclidean distance is a measurement of the straightest and shortest path between two points. For example, the distance between two vectors (p and q) is calculated as the square root of the sum of the squared difference:

$$
d(p, q) = \sqrt{\sum_{i=1}^{n} (p_i - q_i)^2}
$$

This method was already employed in the early publications as a way to evaluate the relationship between metal artefacts and potential ore sources @Stos-Gale.2009 and appear occasionally in recent articles @Westner.2020. Euclidean distance is sometimes a prerequisite step to determine the closeness between data points in cluster analysis. However, critics argue that under this approach both artefact and ore data are examined on a point-by-point basis instead of treated as parts of isotopic populations. In other words, the concepts of artefact assemblage and ore cluster are neglected @Albarede.2020. 

To demonstrate how Euclidean distance can be explored to determine the provenance of artefacts, we compare the distance between a Roman lead ingot and European ore deposits. The smaller the distance is, the closer the ingot matches the reference ore sources. To make the results more intuitive, we adopt a similarity score to measure the closeness @Segaran.2007, as formulated below:

$$
s(p, q) = \frac{1}{1+d(p, q)}
$$

@fig-EDmatching displays a boxplot of the similarity score (= 1 / 1+ED) based on the euclidean distances between one point (Roman ingot) and reference points (ore data). The higher score indicates a higher likelihood to be the source, which should be Northern Eifel in this case.

```{r}
#| label: fig-EDmatching
#| fig-cap: "Similarity of euclidean distance between a Roman ingot and potential mining regions"
#| message: false
#| warning: false

##load dataset
Roman_case <- read.csv("example_dataset/Roman.csv",encoding="UTF-8", stringsAsFactors=FALSE, header = TRUE)

## select ore and ingot data
ore<- Roman_case %>% select(Pb208.Pb204, Pb207.Pb204, Pb206.Pb204, Region_1) %>%
  filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergisches Land",
                         "Laurion", "Cevennes", "Cartagena",
                         "SM-Linares", "Sardinia", "Tuscany"))

ingot<- Roman_case %>% select(Pb208.Pb204, Pb207.Pb204, Pb206.Pb204, Region_1) %>%
  filter(Region_1 %in% c("Roman ingot1")) 

pts <- ingot[, -4]  ## data point
ref <- ore[, -4]    ## reference points
region <- ore$Region_1  ## ore region
distances <- rdist(ref, pts)  ## calculate fields Euclidean Distance

colnames(distances) <- c("distIngot1") 

df <- cbind(ref,region, distances) ## combine region name, raw data, and calculated distances
df1 <- df %>% mutate(similarity_score = 1 / (1 + df$distIngot1))

## plot the boxplot
fig <- plot_ly(df1, y = ~similarity_score, color = ~region,colors = "Dark2", type = "box")
fig
```

### Analytical uncertainty

@Rodriguez.2023 proposes a matching algorithm (AMALIA) based on the analytical uncertainties of isotopic ratios. In order to say that a metal artefact could have been derived from a potential ore deposit, the difference between their lead isotopic compositions should be smaller than, or equal to the respective analytical uncertainty (AU). For any isotope ratio given for two samples A and B to be considered equal, it must fulfill the following equations:


$$
IF (\frac{^{206}Pb}{^{204}Pb})_{A} - (\frac{^{206}Pb}{^{204}Pb})_{B} \le AU_{(\frac{^{206}Pb}{^{204}Pb})_{A,B}},   Then
$$



$$
IF (\frac{^{207}Pb}{^{204}Pb})_{A} - (\frac{^{207}Pb}{^{204}Pb})_{B} \le AU_{(\frac{^{207}Pb}{^{204}Pb})_{A,B}},   Then
$$

$$
IF (\frac{^{208}Pb}{^{204}Pb})_{A} - (\frac{^{208}Pb}{^{204}Pb})_{B} \le AU_{(\frac{^{208}Pb}{^{204}Pb})_{A,B}},   Match (A=B)
$$

The analytical uncertainty can be estimated using the published results for the common lead isotopic standard, NBS-981. @Rodriguez.2023 gathered 358 analyses of NBS-981, measured by either TIMS or MC-ICP-MS, across 88 laboratories. Given that these measurements do not follow a normal distribution, the authors recommended using the relative range (%) to represent the uncertainty, as shown below:

| Summary        | 206Pb/204Pb | 207Pb/208Pb | 208Pb/204Pb | 207Pb/206Pb | 208Pb/206Pb |
|------------|------------|------------|------------|------------|------------|
| Min            | 16.928      | 15.480      | 36.663      | 0.9141      | 2.1648      |
| Max            | 16.945      | 15.502      | 36.736      | 0.9153      | 2.1692      |
| Absolute range | 0.017       | 0.022       | 0.073       | 0.0012      | 0.0044      |
| Relative range | 0.1         | 0.14        | 0.20        | 0.13        | 0.20        |

@tbl-uncertainty shows the differences between a sample data point (a Roman ingot) and reference points (ore data). If this difference is equal to or smaller than the specified analytical uncertainties, the two data points are considered a match. The analysis shows that 81% of ore data from northern Eifel fall within these error ranges, followed by Sauerland with 52%.

```{r}
#| label: tbl-uncertainty
#| tbl-cap: match and mismatch between samples and reference ore data based on analytical uncertainties of SRM-981 published by   Rodríguez et al. 2023 (206Pb/204Pb = 0.1,207Pb/204Pb = 0.14, and 208Pb/204Pb = 0.2). 
#| message: false
#| warning: false

##load dataset
Roman_case <- read.csv("example_dataset/Roman.csv",encoding="UTF-8", stringsAsFactors=FALSE, header = TRUE)

## divide lead isotope data into ore (reference points) and ingot (sample) and extract information about regions  
ore<- Roman_case  %>%
  filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergisches Land",
                         "Laurion", "Cevennes", "Cartagena",
                         "SM-Linares", "Sardinia", "Tuscany"))%>% select(Pb208.Pb204, Pb207.Pb204, Pb206.Pb204)

ingot<- Roman_case  %>%
  filter(Region_1 %in% c("Roman ingot1")) %>% select(Pb208.Pb204, Pb207.Pb204, Pb206.Pb204)

region<- Roman_case  %>% select(Region_1) %>%
  filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergisches Land",
                         "Laurion", "Cevennes", "Cartagena",
                         "SM-Linares", "Sardinia", "Tuscany"))



## the function to check the difference of each ratio between samples and reference points and whether it is equal or smaller 
## than the threshold (analytical uncerntainites)

check_difference <- function(var1, var2, threshold) {
  diff <- var1 - var2
  abs_diff <- abs(diff)
  return(abs_diff <= threshold)
}

## Apply check_difference function to each row in the dataset with defined threshould 0.1, 0.05, 0.1
result_df <- data.frame(
  result_206_204 = apply(ore, 1, function(row) check_difference(ingot$Pb206.Pb204, row["Pb206.Pb204"], 0.1)),
  result_207_204 = apply(ore, 1, function(row) check_difference(ingot$Pb207.Pb204, row["Pb207.Pb204"], 0.14)),
  result_208_204 = apply(ore, 1, function(row) check_difference(ingot$Pb208.Pb204, row["Pb208.Pb204"], 0.2))
)

## Add a column "comparison". If all values are true across three ratios, the column is noted by match otherwise dismatch
result_df$comparison <- ifelse(result_df$result_206_204 & result_df$result_207_204 & result_df$result_208_204, "match", "mismatch")

# attach regions to the result_df
result_df <- cbind(result_df, Region = region$Region_1)

# Group by region and get a summary of match and mismatch
summary_df <- result_df %>%
  group_by(Region) %>%
  summarise(
    match_count = sum(comparison == "match"),
    mismatch_count = sum(comparison == "mismatch"),
    match_percent = match_count / n(),
    mismatch_percent = mismatch_count / n()
  )%>%
  mutate_if(is.numeric, round, 2)

summary_df %>%
  kable(format = "html") %>%
  kable_styling(full_width = FALSE)
```

### Dimension reduction and cluster analysis

Strictly speaking, these statisical techiqnues 
PCA and UPMA, 
HC


### Kernel density estimation

#### Principle

@Baxter.1998 pointed out that lead isotope ratios of many ore deposits actually do not comply with a normal distribution. This prohibits any parametric statistical approach using the mean and standard deviation. Alternatively, both @Beardah.1999 and @Scaife.1999 turned to a more robust non-parametric kernel density estimates (KDEs) to define the isotopic extent of ore fields (as introduced by Chapter 20). It was noted that larger sample sizes were preferable (i.e., several dozen), and that the choice of smoothing and orientation parameters can lead to different KDE distributions @Baxter.1997. The univariate kernel density estimator (\^f h) of the parent function (f) from which the samples are derived is mathematically defined as follows:

$$

\hat{f}_{h}(x) =\frac{1}{n}\sum_{i = 1}^{n} K_{h}(x-x_{i})=\frac{1}{nh}\sum_{i = 1}^{n} K(\frac{x-x_{i}}{h})

$$

where n is the sample size, h is the kernel bandwidth and the kernel K is a summed probability density function over all the samples. According to @Zhang.2006, a commonly used approximation for the optimal bandwidth of the multidimensional kernel is as follows:

$$
h = \{\frac{4}{(d + 2)n}\}^{1/(d+4)}
$$

In this context, d represents the number of dimensions (measured parameters), and n denotes the number of samples in the dataset. The estimated bandwidth ℎ becomes closer to optimal as the data approximate a multivariate normal distribution.

#### Application 1: degree of KDE overlap

The tool [OxPlot](https://c14.arch.ox.ac.uk/oxplot/OxPlot.html) developed by Christopher Bronk Ramsey is capable of testing the degree of overlap between two KDE distributions @Ramsey.2015. In an archaeological sense, we can compare two assemblages X (ore sources) and Y (artefacts) with n and m samples respectively to answer the question "How likely is it that each of the individual samples in assemblage Y could belong to the distribution defined by X?" It can be mathematically expressed as the following equation:

$$
B_{YX} = \frac{\frac{1}{m}\sum_{i = 1}^{n}q_{x}( y_{i})}{\frac{1}{n}\sum_{j = 1}^{n}q_{x\overline{j}}(x_{j})}, B_{XX} = 1
$$

This ratio B<sub>YX</sub> is interpreted as how likely dataset Y is derived from dataset X. This value should approach 1 if the distribution of the two data sets is similar, but it could exceed 1 if dataset Y is located on top of the densest area of data set X. Determining the threshold for the significance of matches is usually arbitrary. Statistically speaking, if the value of the ratio falls below 0.05, then it is probably safe to say, at the equivalent of 95% confidence, that the two distributions are different. This method therefore works best to rule out the potential ore sources.

We treated Roman objects and ore sources as individual assemblages and calculated the overlap between all pairs of KDE distributions. @tbl-kdematrix shows the outcome of a KDE matrix. Based the row G, one can clearly rule out the ore deposits in Cartagena, Laurion, SM-Linares, SM-Los,and Tuscany as likely lead sources due to their insiginifiant likelihoods.  On the contrary, lead ores from Germany (Bergisches Land, Northern Eifel, Sauerland) and France (Montagne Noire) exhibit higher probabilitis and therefore appear to be more relevant with the production of these Roman lead objects. It has to be noted that this approach treats Roman artefacts as a whole assemblage rather than a single object assumed by other approaches.

```{r}
#| label: tbl-kdematrix
#| tbl-cap: KDE overlap matrix between Roman objects and ore data sets 
#| message: false
#| warning: false

##load dataset
kde_matrix <-as.matrix(read.csv("example_dataset/Roman_kde_matrix.csv"))
kde_matrix%>%
  kable(format = "html") %>%
  kable_styling(full_width = FALSE)

```

#### Application 2: KDE probability distributions

@Ceuster.2020 proposed another KDE application to compare a single sample with the probability distributions of targeted mining regions. Since the sum of probability under a normalised density curve is always equal to 1, or 100%, one can compute the probability of a single isotopic measurement within a given interval under that curve. The interval is defined by 10 times the standard deviation (SD) and from which the relative probability for all possible sources is calculated by integrating each density function over that interval. The outcome is displayed as the probability of a single object per region in terms of three commonly-used ratios (206Pb/204Pb, 207Pb/204Pb, and 208Pb/204Pb). The ore deposit containing the highest relative percentage of the all three ratios is considered as the most likely source. The only cavet is that one has to look for the optimal bandwidth for each mining region. At the moment, the authors apply the same bandwidth for all regions. @fig-KDEmatching displays KDE probabilities of a Roman ingot (D-135/1) in comparison with potential mining regions. It clearly shows that there are multiple candidates for the sources including ore deposits in Germany, France and Sardina.

```{r}
#| label: fig-KDEmatching
#| fig-cap: "KDE probabilities of a Roman ingot vesus the potential mining regions. Some regions have near-zero percentages and are not plotted in the diagram"
#| message: false
#| warning: false

##load dataset
Roman_case <- read.csv("example_dataset/Roman.csv",encoding="UTF-8", stringsAsFactors=FALSE, header = TRUE)

## select column names X204.206,207/206,208/206,208/204,207/204,206/204,Region 
data<- Roman_case %>% select(Pb208.Pb204, Pb207.Pb204, Pb206.Pb204, Region_1, Country) %>%
                filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergisches Land",
                         "Laurion", "Cevennes", "Cartagena","Montagne Noire",
                         "SM-Linares", "Sardinia", "Tuscany"))


## select a Roman lead ingot D-135/1 from Bode et al. 2009 (18.332 sd=0.02),(15.605 sd=0.01),(38.272 sd=0.035) 
## and produce the upper and lower bound of 206, 207, 208 using 10xsd. 
## possible improvement: pull out random point from dataset or enter own value
object_206_lower <- 18.332-10*0.02
object_206_upper <- 18.332+10*0.02
object_207_lower <- 15.605-10*0.01
object_207_upper <- 15.605+10*0.01
object_208_lower <- 38.272-10*0.035
object_208_upper <- 38.272+10*0.035

## create an empty dataframe to hold the region and KDE matching results 206/204
result_df_206 <- data.frame(Region_1 = character(0), matching_density = numeric(0)) 

## make a full loop for 206/204
for (region in unique(data$Region_1)) {
  subset_data <- data$Pb206.Pb204[data$Region_1 == region]
  KDE_206 <- density(subset_data,bw=0.05)
  KDE_match_206 <- try(integrate.xy(KDE_206$x,KDE_206$y, object_206_lower, object_206_upper),silent=TRUE)
  if("try-error" %in% class(KDE_match_206)) KDE_match_206 <- 0 #make sure if there is no match, the probability will be zero
  result_df_206 <- rbind(result_df_206, data.frame(Region_1 = region, matching_density = KDE_match_206))
}

## create an empty dataframe to hold the region and KDE matching results 207/204
result_df_207 <- data.frame(Region_1 = character(0), matching_density = numeric(0)) 

## make a full loop for 207/204
for (region in unique(data$Region_1)) {
  subset_data <- data$Pb207.Pb204[data$Region_1 == region]
  KDE_207 <- density(subset_data,bw=0.05)
  KDE_match_207 <- try(integrate.xy(KDE_207$x,KDE_207$y, object_207_lower, object_207_upper),silent=TRUE)
  if("try-error" %in% class(KDE_match_207)) KDE_match_207 <- 0 #make sure if there is no match, the probability will be zero
  result_df_207 <- rbind(result_df_207, data.frame(Region_1 = region, matching_density = KDE_match_207))
}

## create an empty dataframe to hold the region and KDE matching results 208/204
result_df_208 <- data.frame(Region_1 = character(0), matching_density = numeric(0)) 

## make a full loop for 208/204
for (region in unique(data$Region_1)) {
  subset_data <- data$Pb208.Pb204[data$Region_1 == region]
  KDE_208 <- density(subset_data,bw=0.05)
  KDE_match_208 <- try(integrate.xy(KDE_208$x,KDE_208$y, object_208_lower, object_208_upper),silent=TRUE)
  if("try-error" %in% class(KDE_match_208)) KDE_match_208 <- 0 #make sure if there is no match, the probability will be zero
  result_df_208 <- rbind(result_df_208, data.frame(Region_1 = region, matching_density = KDE_match_208))
}


## combine matching results for all isotopic ratios
## combine KDE probabilities of 206,207, and 208 
# table
Region <- result_df_206$Region_1
percentages206 <- result_df_206$matching_density
percentages207 <- result_df_207$matching_density
percentages208 <- result_df_208$matching_density
combine_percentage <-data.frame(percentages206,percentages207,percentages208,Region)


fig <- plot_ly(combine_percentage, x = ~Region, y = ~percentages206, type = 'bar', name = 'percentages206')
fig <- fig %>% add_trace(y = ~percentages207, name = 'percentages207')
fig <- fig %>% add_trace(y = ~percentages208, name = 'percentages208')
fig <- fig %>% layout(yaxis = list(title = ''), barmode = 'group')
fig
```

### Sample convex hull

A typical provenance study tends to define lead isotopic field of an ore deposit/district as a reference to trace the origins of archaeological objects. Nevertheless, @Gentelli.2021 adopted an alternative strategy by grouping sample data to be compared with those of ores. It is achieved through the construction of a "sample convex hull", expressed as the smallest volume in the 3-dimensional space of sample isotope ratios.  They applied this method to the fragments of cut and bent silver items from Iron Age hoards in southern Levant. By clustering samples within each hoard, they identified up too three distinct clusters. The next step was to generate the convex volume bounded by sample data using Delaunay triangulation (Aurenhammer.2013). This avoided the problem of outliers within a sample set that would create the unwanted empty volume in a convex hull. Finally, an algorithm was run for each data point of the ore database against each convex hull. Each ore data falling inside the hull of each hoard are considered a match. 

### Binary mixing model

To address the extent of mixing and recycling in ancient Greek silver coinage, Albarede et al. (2024) turned to the distribution of lead isotope data and came up with a new binary mixing model to estimate the production of fresh and scrap silver for minting. At the first step, the authors performed the PCA on 112 silver coins from Athens and examined which principal component would explain the variability of data. It turned out that the first component c1 accounts for 97.7% of variance with the rather small contribution from c2 and c3. This indicates that no more than two end-members are present in the mixture.



### Bayesian mixing model

## Self check

Now you can provide answers to the following questions:

-   ...

## Further reading

To delve deeper into this topic, we recommend the following resources:

-   Pollard, A. M., and Bray, P. J., 2015, A New Method For Combining Lead Isotope and Lead Abundance Data to Characterize Archaeological Copper Alloys\*: Combining lead isotope and lead abundance data, Archaeometry, 57(6), 996–1008.
-   De Ceuster, S., and Degryse, P., 2020, A ‘match–no match‘ numerical and graphical kernel density approach to interpreting lead isotope signatures of ancient artefacts, Archaeometry, 62(S1), 107–16.
-   Tomczyk, C., and Żabiński, G., 2023, A PCA-AHC Approach to Provenance Studies of Non-Ferrous Metals with Combined Pb Isotope and Chemistry Data, Journal of Archaeological Method and Theory, 1–51.
-   Gentelli, L., Blichert-Toft, J., Davis, G., Gitler, H., and Albarède, F., 2021, Metal provenance of Iron Age Hacksilber hoards in the southern Levant, Journal of Archaeological Science, 134, 105472.
-   Hsu, Y.-K., Rawson, J., Pollard, A. M., Ma, Q., Luo, F., Yao, P.-H., and Shen, C.-C., 2018, Application of Kernel Density Estimates to Lead Isotope Compositions of Bronzes from Ningxia, North-West China: Application of KDEs to lead isotope compositions, Archaeometry, 60(1), 128–43.
-   Birch, T., Westner, K. J., Kemmers, F., Klein, S., Höfer, H. E., and Seitz, H. ‐M., 2019, Retracing Magna Graecia’s silver: coupling lead isotopes with a multi‐standard trace element procedure, Archaeometry, arcm.12499.
-   Delile, H., Blichert-Toft, J., Goiran, J.-P., Keay, S., and Albarède, F., 2014, Lead in ancient Rome’s city waters, Proceedings of the National Academy of Sciences, 111(18), 6594–9.
-   Baxter, M. J., Beardah, C. C., and Westwood, S., 2000, Sample size and related Issues in the Analysis of lead Isotope data, Journal of Archaeological Science, 27(10), 973–80.
-   Birch, T., Kemmers, F., Klein, S., Seitz, H. and Höfer, H., 2020. Silver for the Greek colonies: Issues, analysis and preliminary results from a large‐scale coin sampling project. Metallurgy in numismatics, 6, pp.101-48.
-   Sayre, E.V., Yener, K.A., Joel, E.C. and Barnes, I.L., 1992. Statistical evaluation of the presently accumulated lead isotope data from Anatolia and surrounding regions. Archaeometry, 34(1), pp.73-105.
-   Stos-Gale, Z.A. and Gale, N.H., 2009. Metal provenancing using isotopes and the Oxford archaeological lead isotope database (OXALID). Archaeological and Anthropological Sciences, 1, pp.195-213.
-   Bode, M., Hauptmann, A. and Mezger, K. 2009. Tracing Roman lead sources using lead isotope analyses in conjunction with archaeological and epigraphic evidence—a case study from Augustan/Tiberian Germania. Archaeol Anthropol Sci 1, 177–194 (2009). https://doi.org/10.1007/s12520-009-0017-0
-   Rodríguez, J., Sinner, A.G., Martínez-Chico, D. and Zalduegui, J.F.S., 2023. AMALIA, A Matching Algorithm for Lead Isotope Analyses: Formulation and proof of concept at the roman foundry of Fuente Spitz (Jaén, Spain). Journal of Archaeological Science: Reports, 51, p.104192.
