---
title: Matching of Pb isotope data 
author: 
  - name: Yiu-Kang Hsu
    orcid: 0000-0002-2439-4863
    corresponding: true
    email: Yiu-Kang.Hsu@bergbaumuseum.de
---

## Learning objective

::: {style="text-align: justify"}
In this chapter, we will employ various statistical methods to compare lead isotope ratios between archaeological samples and reference ore data. In principle, provenance studies of lead isotopes aim to exclude unlikely ore signatures to narrow down potential sources. This "exclusion exercise" often involves defining ore fields, discriminating between possible ore sources, and estimating the influence of recycling and mixing when provenance inquiries are inconclusive.
:::

## Prior knowledge

::: {style="text-align: justify"}
It is assumed that readers are familiar with concepts of the previous chapter on data visualization and possess a basic understanding of R programming. In this chapter, we will utilize a suite of R packages to perform statistical assignments, allowing for a quantitative exploration of isotopic matches or mismatches between artifacts and ores.
:::

## Outline

::: {style="text-align: justify"}
The "eyeballing" technique is the simplest and most straightforward method for interpreting lead isotopic ratios. However, it can become unreliable and subjective under certain circumstances, especially when it comes to large and heterogeneous datasets. Employing statistical tools offers a more objective and effective way to analyze complex data patterns. The following section will introduce various statistical methods recently applied to the study of lead isotopes in archaeological metals. We use the dataset of Roman lead objects published by Bode et al. (2009) to showcase these different approaches.
:::

The following statistical methods are discussed: 

<ul>
<li>Euclidean distance</li>
<li>Analytical uncertainties</li>
<li>Dimensional reduction and cluster analysis</li>
<li>Kernel density estimation</li>
<li>Convex hull</li>
<li>Binary mixing model</li>
<li>Bayesian mixing model</li>
</ul>

## Learning content

Let's start with loading the required R packages:

```{r}
#| label: load-pkgs
#| code-summary: "Packages"
#| message: false
#| warning: false

library(plotly) # for interactive plots
library(ggtern) # for ternary kernel density plots
library(downlit)# code linking
library(xml2)   # code linking
library(dplyr)  # for tiding up data sets
library(sfsmisc) # kde probabilities 
library(fields)  # calculation of distance between points
library(kableExtra) # for coloring fields of a table
```

### Euclidean distance

Euclidean distance is a measurement of the straightest and shortest path between two points. For example, the distance between two vectors (p and q) is calculated as the square root of the sum of the squared difference:

$$
d(p, q) = \sqrt{\sum_{i=1}^{n} (p_i - q_i)^2}
$$

The method was already employed in the early publications as a way to evaluate the relationship between metal artefacts and potential ore sources (e.g. Stos-Gale and Gale 2009) and appear occasionally in recent articles (e.g. Westner et al. 2020). For the cluster analysis, Euclidean distance is sometimes a prerequisite step to determine the closeness between data points. However, critics argue that under this approach both artifact and ore data are examined on a point-by-point basis instead of treated as parts of isotopic populations. In other words, the concepts of artifact assemblage and ore cluster are neglected (Albarede et al. 2020)

To demonstrate how Euclidean distance can be explored to determine the provenance of artifacts, we compare the distance between a Roman lead ingot and European ore deposits. The smaller the distance is, the closer the ingot matches the reference ore sources. To make the results more intuitive, we adopt a similarity score (s) proposed by Toby Segaran (Segaran 2007), as formulated below: 

$$
s(p, q) = \frac{1}{1+d(p, q)}
$$




@fig-EDmatching displays a boxplot of the similarity score (= 1 / 1+ED) based on the euclidean distances between one point (Roman ingot) and reference points (ore data). The higher score indicates a higher likelihood to be the source, which should be Northern Eifel in this case. 

```{r}
#| label: fig-EDmatching
#| fig-cap: "Similarity of euclidean distance between a Roman ingot and potential mining regions"
#| message: false
#| warning: false

##load dataset
Roman_case <- read.csv("example_dataset/Roman_1.csv",encoding="UTF-8", stringsAsFactors=FALSE, header = TRUE)

## select ore and ingot data
ore<- Roman_case %>% select(X208.204, X207.204, X206.204, Region_1) %>%
  filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergische Land",
                         "Laurion", "Cevennes", "Cartagena",
                         "SM-Linares", "Sardinia", "Tuscany"))

ingot<- Roman_case %>% select(X208.204, X207.204, X206.204, Region_1) %>%
  filter(Region_1 %in% c("Roman ingot1")) 

pts <- ingot[, -4]  ## data point
ref <- ore[, -4]    ## reference points
region <- ore$Region_1  ## ore region
distances <- rdist(ref, pts)  ## calculate fields Euclidean Distance

colnames(distances) <- c("distIngot1") 

df <- cbind(ref,region, distances) ## combine region name, raw data, and calculated distances
df1 <- df %>% mutate(similarity_score = 1 / (1 + df$distIngot1))

## plot the boxplot
fig <- plot_ly(df1, y = ~similarity_score, color = ~region,colors = "Dark2", type = "box")
fig
```

### Analytical uncertainty

Rodríguez et al. 2023 propose "A Matching Algorithm for Lead Isotope Analyses" (AMALIA) based on the analytical uncertainties of isotopic ratios. In order to say that a metal artifact could have been derived from a pontential ore deposit, the difference between their lead isotopic compositions should be smaller than, or equal to the respective analytical uncertainty (AU). For any isotope ratio given for two samples a and b to be considered equal, it must fulfill the following equaitons: 


$$
IF (\frac{^{206}Pb}{^{204}Pb})_{a} - (\frac{^{206}Pb}{^{204}Pb})_{b} \le AU_{(\frac{^{206}Pb}{^{204}Pb})_{a,b}},   Then
$$
$$
IF (\frac{^{207}Pb}{^{204}Pb})_{a} - (\frac{^{207}Pb}{^{204}Pb})_{b} \le AU_{(\frac{^{207}Pb}{^{204}Pb})_{a,b}},   Then
$$
$$
IF (\frac{^{208}Pb}{^{204}Pb})_{a} - (\frac{^{208}Pb}{^{204}Pb})_{b} \le AU_{(\frac{^{208}Pb}{^{204}Pb})_{a,b}},   Match (a=b)
$$

The anaytical uncertainty can be estimated from the published results of the 
Given that published LIA are obtained at different laboratories and in order to compare them, the analytical uncertainty has to consider the variety of results that can be obtained at any laboratory for the same sample. This can be estimated from the published results of the interlaboratory lead standard known as NBS-981 (National Institute of Standards and Technology -formerly National Bureau of Standards-, USA). Rodríguez et al. (2023) collected 358 analyses of NBS-981 datast measured by TIMS or MC-ICP-MS obtain LIA in 88 laboratories. Due to non-normal distribution of these measurements, the authors suggested the adoption of the relative range (%) of data spread from the minumun to the maxiumum ratios, as shown below: 

| Summary       | 206Pb/204Pb | 207Pb/208Pb | 208Pb/204Pb | 207Pb/206Pb | 208Pb/206Pb |
| -----------   | ----------- | ----------- | ----------- | ----------- | ----------- |
|Min            | 16.928      | 15.480      | 36.663      | 0.9141      | 2.1648      |
|Max            | 16.945      | 15.502      | 36.736      | 0.9153      | 2.1692      |
|Absolute range | 0.017       | 0.022       | 0.073       | 0.0012      | 0.0044      |
|Relative range | 0.1         | 0.14        | 0.20        | 0.13        | 0.20        |




@tbl-uncertainty show the difference between a sample point (Roman ingot) and reference points (ore data). If the difference is equal or smaller then defined analytical uncertainties, two data points are considered to be a match. This approach is introduced by Rodríguez et al. (2023)

```{r}
#| label: tbl-uncertainty
#| tbl-cap: match and mismatch between samples and reference ore data based on analytical uncertainties of SRM-981 published by   Rodríguez et al. 2023 (206Pb/204Pb = 0.1,207Pb/204Pb = 0.14, and 208Pb/204Pb = 0.2). 
#| message: false
#| warning: false

##load dataset
Roman_case <- read.csv("example_dataset/Roman_1.csv",encoding="UTF-8", stringsAsFactors=FALSE, header = TRUE)

## divide lead isotope data into ore (reference points) and ingot (sample) and extract information about regions  
ore<- Roman_case  %>%
  filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergische Land",
                         "Laurion", "Cevennes", "Cartagena",
                         "SM-Linares", "Sardinia", "Tuscany"))%>% select(X208.204, X207.204, X206.204)

ingot<- Roman_case  %>%
  filter(Region_1 %in% c("Roman ingot1")) %>% select(X208.204, X207.204, X206.204)

region<- Roman_case  %>% select(Region_1) %>%
  filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergische Land",
                         "Laurion", "Cevennes", "Cartagena",
                         "SM-Linares", "Sardinia", "Tuscany"))



## the function to check the difference of each ratio between samples and reference points and whether it is equal or smaller 
## than the threshold (analytical uncerntainites)

check_difference <- function(var1, var2, threshold) {
  diff <- var1 - var2
  abs_diff <- abs(diff)
  return(abs_diff <= threshold)
}

## Apply check_difference function to each row in the dataset with defined threshould 0.1, 0.05, 0.1
result_df <- data.frame(
  result_206_204 = apply(ore, 1, function(row) check_difference(ingot$X206.204, row["X206.204"], 0.1)),
  result_207_204 = apply(ore, 1, function(row) check_difference(ingot$X207.204, row["X207.204"], 0.14)),
  result_208_204 = apply(ore, 1, function(row) check_difference(ingot$X208.204, row["X208.204"], 0.2))
)

## Add a column "comparison". If all values are true across three ratios, the column is noted by match otherwise dismatch
result_df$comparison <- ifelse(result_df$result_206_204 & result_df$result_207_204 & result_df$result_208_204, "match", "mismatch")

# attach regions to the result_df
result_df <- cbind(result_df, Region = region$Region_1)

# Group by region and get a summary of match and mismatch
summary_df <- result_df %>%
  group_by(Region) %>%
  summarise(
    match_count = sum(comparison == "match"),
    mismatch_count = sum(comparison == "mismatch"),
    match_percent = match_count / n(),
    mismatch_percent = mismatch_count / n()
  )%>%
  mutate_if(is.numeric, round, 2)

summary_df %>%
  kable(format = "html") %>%
  kable_styling(full_width = FALSE)
```

@fig-KDEmatching displays KDE probabilities of a sample derived from potential mining regionsregions

```{r}
#| label: fig-KDEmatching
#| fig-cap: "KDE probabilities of the potential mining regions"
#| message: false
#| warning: false

##load dataset
Roman_case <- read.csv("example_dataset/Roman_1.csv",encoding="UTF-8", stringsAsFactors=FALSE, header = TRUE)

## select column names X204.206,207/206,208/206,208/204,207/204,206/204,Region 
data<- Roman_case %>% select(X208.204, X207.204, X206.204, Region_1, Country) %>%
                filter(Region_1 %in% c("Northern Eifel", "Sauerland", "Bergische Land",
                         "Laurion", "Cevennes", "Cartagena",
                         "SM-Linares", "Sardinia", "Tuscany"))


## select  random Roman lead object D-121/1 from Bode et al. 2009 (18.412 sd=0.02),(15.657 sd=0.01),(38.497 sd=0.035) 
## and produce the upper and lower bound of 206, 207, 208 using 10xsd. 
## possible improvement: pull out random point from dataset or enter own value
object_206_lower <- 18.412-10*0.02
object_206_upper <- 18.412+10*0.02
object_207_lower <- 15.657-10*0.01
object_207_upper <- 15.657+10*0.01
object_208_lower <- 38.497-10*0.035
object_208_upper <- 38.497+10*0.035

## create an empty dataframe to hold the region and KDE matching results 206/204
result_df_206 <- data.frame(Region_1 = character(0), matching_density = numeric(0)) 

## make a full loop for 206/204
for (region in unique(data$Region_1)) {
  subset_data <- data$X206.204[data$Region_1 == region]
  KDE_206 <- density(subset_data,bw=0.03)
  KDE_match_206 <- try(integrate.xy(KDE_206$x,KDE_206$y, object_206_lower, object_206_upper),silent=TRUE)
  if("try-error" %in% class(KDE_match_206)) KDE_match_206 <- 0 #make sure if there is no match, the probability will be zero
  result_df_206 <- rbind(result_df_206, data.frame(Region_1 = region, matching_density = KDE_match_206))
}

## create an empty dataframe to hold the region and KDE matching results 207/204
result_df_207 <- data.frame(Region_1 = character(0), matching_density = numeric(0)) 

## make a full loop for 207/204
for (region in unique(data$Region_1)) {
  subset_data <- data$X207.204[data$Region_1 == region]
  KDE_207 <- density(subset_data,bw=0.03)
  KDE_match_207 <- try(integrate.xy(KDE_207$x,KDE_207$y, object_207_lower, object_207_upper),silent=TRUE)
  if("try-error" %in% class(KDE_match_207)) KDE_match_207 <- 0 #make sure if there is no match, the probability will be zero
  result_df_207 <- rbind(result_df_207, data.frame(Region_1 = region, matching_density = KDE_match_207))
}

## create an empty dataframe to hold the region and KDE matching results 208/204
result_df_208 <- data.frame(Region_1 = character(0), matching_density = numeric(0)) 

## make a full loop for 208/204
for (region in unique(data$Region_1)) {
  subset_data <- data$X208.204[data$Region_1 == region]
  KDE_208 <- density(subset_data,bw=0.03)
  KDE_match_208 <- try(integrate.xy(KDE_208$x,KDE_208$y, object_208_lower, object_208_upper),silent=TRUE)
  if("try-error" %in% class(KDE_match_208)) KDE_match_208 <- 0 #make sure if there is no match, the probability will be zero
  result_df_208 <- rbind(result_df_208, data.frame(Region_1 = region, matching_density = KDE_match_208))
}


## combine matching results for all isotopic ratios
## combine KDE probabilities of 206,207, and 208 
# table
Region <- result_df_206$Region_1
percentages206 <- result_df_206$matching_density
percentages207 <- result_df_207$matching_density
percentages208 <- result_df_208$matching_density
combine_percentage <-data.frame(percentages206,percentages207,percentages208,Region)


fig <- plot_ly(combine_percentage, x = ~Region, y = ~percentages206, type = 'bar', name = 'percentages206')
fig <- fig %>% add_trace(y = ~percentages207, name = 'percentages207')
fig <- fig %>% add_trace(y = ~percentages208, name = 'percentages208')
fig <- fig %>% layout(yaxis = list(title = '%'), barmode = 'group')
fig
```

## Self check

Now you can provide answers to the following questions:

-   ...

## Further reading

To delve deeper into this topic, we recommend the following resources...

-   Pollard, A. M., and Bray, P. J., 2015, A New Method For Combining Lead Isotope and Lead Abundance Data to Characterize Archaeological Copper Alloys\*: Combining lead isotope and lead abundance data, Archaeometry, 57(6), 996–1008.
-   De Ceuster, S., and Degryse, P., 2020, A ‘match–no match‘ numerical and graphical kernel density approach to interpreting lead isotope signatures of ancient artefacts, Archaeometry, 62(S1), 107–16.
-   Tomczyk, C., and Żabiński, G., 2023, A PCA-AHC Approach to Provenance Studies of Non-Ferrous Metals with Combined Pb Isotope and Chemistry Data, Journal of Archaeological Method and Theory, 1–51.
-   Gentelli, L., Blichert-Toft, J., Davis, G., Gitler, H., and Albarède, F., 2021, Metal provenance of Iron Age Hacksilber hoards in the southern Levant, Journal of Archaeological Science, 134, 105472.
-   Hsu, Y.-K., Rawson, J., Pollard, A. M., Ma, Q., Luo, F., Yao, P.-H., and Shen, C.-C., 2018, Application of Kernel Density Estimates to Lead Isotope Compositions of Bronzes from Ningxia, North-West China: Application of KDEs to lead isotope compositions, Archaeometry, 60(1), 128–43.
-   Birch, T., Westner, K. J., Kemmers, F., Klein, S., Höfer, H. E., and Seitz, H. ‐M., 2019, Retracing Magna Graecia’s silver: coupling lead isotopes with a multi‐standard trace element procedure, Archaeometry, arcm.12499.
-   Delile, H., Blichert-Toft, J., Goiran, J.-P., Keay, S., and Albarède, F., 2014, Lead in ancient Rome’s city waters, Proceedings of the National Academy of Sciences, 111(18), 6594–9.
-   Baxter, M. J., Beardah, C. C., and Westwood, S., 2000, Sample Size and Related Issues in the Analysis of Lead Isotope Data, Journal of Archaeological Science, 27(10), 973–80.
